---
title: "Luck versus Skills. Is Your Assessment of Sales Reps' Performance Truly Unbiased?"
author: "Daniel Szulc"
date: "8 September 2015"
output: html_document
---
```{r initializeOptions, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE, cache = TRUE,warning = FALSE, message = FALSE)
```

```{r loadingLibraries}
library(dplyr)
library(tidyr)
library(Hmisc)

```

```{r loadingData}
if (!file.exists("./2015-09-08-Luck-versus-Skills-Is-Your-Assessment-of-Sales-Reps-Performance-Truly-Unbiased_files/salesPerformanceData.rds")) {
       download.file(url = "http://danielszulc.intelligence-data.net/2015-09-08-Luck-versus-Skills-Is-Your-Assessment-of-Sales-Reps-Performance-Truly-Unbiased_files/salesPerformanceData.rds", destfile = "./2015-09-08-Luck-versus-Skills-Is-Your-Assessment-of-Sales-Reps-Performance-Truly-Unbiased_files/salesPerformanceData.rds")
}
salesData <- readRDS("./2015-09-08-Luck-versus-Skills-Is-Your-Assessment-of-Sales-Reps-Performance-Truly-Unbiased_files/salesPerformanceData.rds")
```
Sales reps are probably the most frequent evaluated employees in companies. Monthly or quarterly quotas they have to reach are natural opportunities to rate performance and give some feedback. It seems that assessment of such employees is quite straightforward and objective. Unlike in cases of many other types of jobs, here we have quantitative measures that reflect performance. We can count how much he or she sold. We can express this number as revenue in dollars. We can count a number of new customers or any other number relevant to the particular industry or company. We can quickly evaluate if a sales rep achieved a quota or if he sold more or less than his peers. However is this truly as easy and straightforward as it seems?

## Outcome Bias

In 1988, Jonathan Baron and John C. Hershey wrote a fascinating article[^2] describing in details a phenomenon called outcome bias. Despite it was almost 30 years ago, their observations and conclusions seem to be still valid. In short, people tend to rate other people's actions, decisions and competencies better when the outcome of their activities is favourable. For example, when we observe two people playing in a pure random game, we usually assess the winner as a better player even if we know the rules and understand that this is a 100% chance game. Things become harder when the situation we have to judge is a mixture of skills and luck. And that is the case in the most real life problems. Getting back to our sales reps evaluation problem, we should be aware that the outcome, in this situation, is a rather complex effect of many different factors. Some of them fully depend on an employee. Others are totally out of sales rep's control. So she can decide on what and how she does, how she react to certain circumstances. This is what we call skill element. However, she can hardly influence the general market sentiment, macroeconomic situation, actions taken by competitors or customers' internal decision processes. All of these are to some extent random variables and represent something we can call luck or chance. When you focus primarily on the outcome (i.e. value or quantity sold), you can overestimate the influence of these random factors and underestimates real skill involved in the process of achieving the targets. In fact, a job of sales rep may be perceived as one of the most affected by the randomness of an environment since it requires interpersonal interactions and deals with competitive forces. Excellent result of a rep in a given period may be (and to some extent certainly is) a product of favourable external conditions. A weak result in the following quarter almost always involves unfavourable circumstances. Of course, there are cases (and surely every sales manager knows them well) when this kind of explanation is used as an excuse by slacking employees. This makes the problem of appropriate performance evaluation even more important.

## Regression to the Mean and Separation Skills from Luck

We would like to be able to say how much of performance variability (peaks and holes of sales from period to period, differences in outcomes of reps) can be explained by chance factors and how much is truly determined by underlying differences in skills and effort. To reach this goal we can use a concept called regression to the mean. Many authors observed and described this phenomenon in the context of various industries and situations. 
J. M. Bland and D. G. Altman in their short article[^3] give a very concise and not-so-mathematical explanation of the concept. David Lane gives a few illustrative examples too[^1]. Generally speaking, the rule is that in the long run, outstanding observations tend to back to their usual, normal levels. If the sales performance is a matter of chance, an excellent result of a particularly successful rep in a given period should deteriorate in the following months. The shorter the periods, the stronger the effect we can expect. Skill elements should be persistent and change much less from month to month. To illustrate this phenomenon on a real example from the world of sales performance analysis, I would like to share some thoughts on an observational study I was involved in. We conducted our study during a project in one of the financial services institutions in Poland. We analysed its sales reps performance. Their work concentrated on a business-to-customer segment. They were selling various financial products to individuals, including also microentrepreneurs. It is worth to note that the job of our subjects involved tasks characteristics of relationship bankers. However, an important part of their daily duties was prospecting via outbound cold calls. They were processing and closing deals as well. Due to confidentiality reasons we are not able to disclose any further details regarding this company and its employees. Moreover, we had to obfuscate some data (by applying changes of scales and other techniques) to avoid any unintended disclosure. Nevertheless, all the observations based on this data are still genuinely valid. In the first month of the study, we ranked the sales representatives. The ranking based on a percentage of the sales quota achieved. We repeated it for the two subsequent months. The below graph illustrates the results.

```{r fig1}
# prepare data for a plot
salesRanking <- salesData %>% group_by(month) %>% 
        select(obfEmail,month, newCustomersSales) %>% 
        arrange(desc(newCustomersSales)) %>%
        mutate(ranking = min_rank(-newCustomersSales)) %>% arrange(month, ranking)

salesRankingCompare <- salesRanking %>% select(-newCustomersSales) %>% spread(key = month, value = ranking)
colnames(salesRankingCompare) <-make.names(colnames(salesRankingCompare)) 

salesRankingCompare <- salesRankingCompare[complete.cases(salesRankingCompare),]

# create a plot
plot(x=rep(1,nrow(salesRankingCompare)), y=salesRankingCompare$X1, xlim=c(1,3), ylim=rev(c(1,160)),xaxt="n", main = "Changes in monthly sales ranking", ylab="Position in sales ranking",xlab="")
axis(side = 1,at = c(1,2,3),labels = c("Feb", "Mar", "Apr"))
points(x=rep(2,nrow(salesRankingCompare)), y=salesRankingCompare$X2)
segments(x0=rep(1,nrow(salesRankingCompare)),y0=salesRankingCompare$X1,x1=rep(2,nrow(salesRankingCompare)), y1=salesRankingCompare$X2, lty=3)
points(x=rep(3,nrow(salesRankingCompare)), y=salesRankingCompare$X3)
segments(x1=rep(3,nrow(salesRankingCompare)),y1=salesRankingCompare$X3,x0=rep(2,nrow(salesRankingCompare)), y0=salesRankingCompare$X2, lty=3)
```

Please note that we included only employees that were selling in each month. Hence, there are some gaps in the plot. The number of active reps varied due to staff turnover and other reasons (for example holidays, sick leaves, internal promotions). It seems like there are sharp fluctuations in sales reps' ranking. An employee in the top 10 of one month drops to the bottom half of the ranking. On the other hand, underperformers from bottom 100 can progress to top 50 in a matter of a month. To draw a full picture, we should also note that some top performers can hold their high positions. Nevertheless, the correlation between positions in the ranking from month to month is in the best case moderate. In most cases it is rather weak. It can be clearly seen in the below table.

```{r calculateCorrelations}
corTable <-rcorr(as.matrix(salesRankingCompare[,c(2:4)]), type="spearman")
corTable <-corTable$r

colnames(corTable) <- c("Feb","Mar", "Apr")
rownames(corTable) <- c("Feb","Mar", "Apr")
```

```{r tableShow}
knitr::kable(corTable,digits = 2,caption = "Table: Spearman's Rang Correlation Coefficients for Sales Rankings",)
```

Based on these data we can make two points.

1. It is a rather hard task to predict the performance of a particular sales rep based on her sales outcome from the previous month.  

2. It can be quite dangerous and costly to assess the performance of sales reps based solely on the outcome numbers as it can lead to serious misjudgments. If we limit our evaluation only to a result, our assessment can vary dramatically from month to month. The ratings become inconsistent and of little informative value for the personnel.

### Focus on Process

The outcome bias leads to misjudgment and potentially wrong decisions. Hence, we should try to shift our focus from the pure results-based measures to the indicators that reflect how the sales process is conducted. We can collect data that describe employee's behavior and shed some light on how she is driving toward sales outcomes. It can help to deeper understand the sales process and the meaning of different intermediate steps. The steps that need to be taken between generating a lead and closing a deal. We can capture data that show the efficiency of different steps (e.g. cold calling) in the sales process. We can analyse the activity of sales reps and their time management. In our example of the financial institution, we identified an average number of cold calls made daily and a number of hours spend on selling activities as more stable and predictive variables. There is no space here to describe in details these discoveries (you can read about them here). There may also be no point in it as we can be pretty sure that in different environments, industries, organizational culture, different time and space, you have to find and analyse different metrics. Nevertheless, there are numerous tools that can help you in this process. You can find many descriptions of suggested sales metrics. I recommend the article published by insightsquared.com website[^is] as a good starting point. This metrics can be generally classified into three categories: activity metrics, outcome metrics and mixed metrics (for example various ratios of activity and outcome variables). Choose a bunch of them that seem appropriate in your environment. Then you can use more or less advanced statistical data analysis tools to find correlations and build an explanatory model. As a starting point here, you can use the old good linear regression.  

Just to be clear, I do not suggest abandoning outcome-based metrics in sales performance evaluation. Whatever we say, at the end of the day it is important how much we sold. However, paying attention also to process-based metrics can give every sales manager valuable insights and help to separate skill from chance and let assess the sales reps in a more objective way.

#### Additional Materials
[Reproducible R scrpit](http:/danielszulc.intelligence-data.net/2015-09-08-Luck-versus-Skills-Is-Your-Assessment-of-Sales-Reps-Performance-Truly-Unbiased.Rmd) 

[Dataset used in the post](http://danielszulc.intelligence-data.net/2015-09-08-Luck-versus-Skills-Is-Your-Assessment-of-Sales-Reps-Performance-Truly-Unbiased_files/salesPerformanceData.rds)  

#### References and Sources

[^1]: D.M. Lane, HyperStat Online Statistics Textbook, Accesed 28 August 2015, http://davidmlane.com/hyperstat/B153351.html.

[^2]: J. Baron and J. C.Hershey, "Outcome bias in decision evaluation",  *Journal of Personality and Social Psychology* 54 (1988): 569-579. Accessed 27 August 2015, https://www.sas.upenn.edu/~baron/papers.htm/judg.html.

[^3]: J. M. Bland and D. G. Altman, "Statistic Notes: Regression towards the mean", BMJ 308 (1994): 1499, doi: http://dx.doi.org/10.1136/bmj.308.6942.1499.
[^is]: Zorian Rotenberg, "The Right KPIs",insightsquared.com, 24 May 2013, accessed 1 September 2015, http://www.insightsquared.com/2013/05/the-right-key-sales-performance-metrics-for-your-sales-reps/.   
